---
title: "blastn_tax_analysis"
author: "Kimberly Ledger"
date: "2023-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
```

```{r load-output-from-taxonkit}
#taxonomy <-read.delim("/genetics/edna/workdir/gadids/20230111/S1_ND1_529_789/trimmed/filtered/outputs/blastn_taxlineage.txt", header = FALSE, na.strings=c(""," ","NA"))

#taxonomy <-read.delim("/genetics/edna/workdir/gadids/20230111/S2_ND1_651_862/trimmed/filtered/outputs/blastn_taxlineage.txt", header = FALSE, na.strings=c(""," ","NA"))

#taxonomy <-read.delim("/genetics/edna/workdir/gadids/20230111/S3_ND5_547_800/trimmed/filtered/outputs/blastn_taxlineage.txt", header = FALSE, na.strings=c(""," ","NA"))

#taxonomy <-read.delim("/genetics/edna/workdir/gadids/20230111/S4_COX3_422v1_598/trimmed/filtered/outputs/blastn_taxlineage.txt", header = FALSE, na.strings=c(""," ","NA"))

#taxonomy <-read.delim("/genetics/edna/workdir/gadids/20230111/S5_COX3_422v2_598/trimmed/filtered/outputs/blastn_taxlineage.txt", header = FALSE, na.strings=c(""," ","NA"))

#taxonomy <-read.delim("/genetics/edna/workdir/gadids/20230111/S6_ND5_1036_1285/trimmed/filtered/outputs/blastn_taxlineage.txt", header = FALSE, na.strings=c(""," ","NA"))

#taxonomy <-read.delim("/genetics/edna/workdir/bristolbay/trimmed/filtered_kl/outputs/blastn_taxlineage.txt", header = FALSE, na.strings=c(""," ","NA"))

#taxonomy <- read.delim("/genetics/edna/workdir/gadids/20230111/S1_ND1_529_789/trimmed/filtered/outputs/test2_blastn_taxlineage.txt", header = FALSE, na.strings=c(""," ","NA"))

head(taxonomy)
```


# clean up the header a bit
```{r}
# use the full taxonomy rather than the seq id to collapse identical entries
tax_df <- taxonomy %>%
  filter(V4 > 100) %>% # make sure all retained matches are >100 bp
  select(-V2, -V5, -V6, -V7, -V14, -V8, -V9, -V10, -V11, -V12, -V15, -V16) %>%  #remove unnecessary columns
  group_by(V1, V17) %>% # group by the sequence key and the full taxonomy to reduce duplicate entries
  unique() %>% # doing that reduced the number of entries from 146k to 17k
  rename(qseqid=V1, perc_id=V3, length=V4, sscinames=V13, taxonomy=V17) %>% #rename headers
  filter(!str_detect(sscinames, "environmental")) %>% # filter out any environmental samples
  filter(!str_detect(sscinames, "synthetic")) #%>% # filter out any synthetic "samples"
  #filter(perc_id >= 98) # seems like some of the matches below 98% are dubious (jellyfish and herring <1% different??)
```

```{r how-many-total-ASVs-w-taxonomy}
tax_df %>% 
  ungroup() %>%
  select(qseqid) %>%
  unique() %>%
  nrow
```

# reformat table 

```{r}
# formatting the taxonomy variables
taxon_df <- tax_df %>%
  filter(str_detect(taxonomy, ";")) %>%
  separate(taxonomy, into=c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = ";")
```


## Sort and clean-up the results based on % identity

```{r}
# 1. rename Gadus ogac to Gadus macrocephalus

taxon_df$species[taxon_df$species == "Gadus ogac"] <- "Gadus macrocephalus"
```

```{r}
taxon_df <- taxon_df %>%
  select(qseqid, species) %>%
  group_by(qseqid, species) %>%
  unique()
```

do any ASVs have more than one species-level id? 
```{r}
taxon_df[duplicated(taxon_df$qseqid),]
```

## create final taxonomy table for output
```{r}
taxonomy <- taxon_df %>%
  rename(ASV = qseqid) %>%
  rename(taxon = species)

taxonomy$taxonomic_level <- "species"
```


```{r}
#write.csv(taxonomy, "/genetics/edna/workdir/gadids/20230111/S1_ND1_529_789/trimmed/filtered/outputs/asv_taxonomy_blastn.csv")
#write.csv(taxonomy, "/genetics/edna/workdir/gadids/20230111/S2_ND1_651_862/trimmed/filtered/outputs/asv_taxonomy_blastn.csv")
#write.csv(taxonomy, "/genetics/edna/workdir/gadids/20230111/S3_ND5_547_800/trimmed/filtered/outputs/asv_taxonomy_blastn.csv")
#write.csv(taxonomy, "/genetics/edna/workdir/gadids/20230111/S4_COX3_422v1_598/trimmed/filtered/outputs/asv_taxonomy_blastn.csv")
#write.csv(taxonomy, "/genetics/edna/workdir/gadids/20230111/S5_COX3_422v2_598/trimmed/filtered/outputs/asv_taxonomy_blastn.csv")
#write.csv(taxonomy, "/genetics/edna/workdir/gadids/20230111/S6_ND5_1036_1285/trimmed/filtered/outputs/asv_taxonomy_blastn.csv")
```

did not output asv taxonomy for S6 because multiple species of gadids were included in the same ASV... i.e. multiple species have >96% similarity for some ASVs.  i could go back and re run the blastn with more stringent perc_identity cut off. not doing this for now because S5 was not included in mock community tests. 


