---
title: "taxonomic analysis of eDNA samples"
author: "Kimberly Ledger"
date: "2022-11-08"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

this script using the insect R package for taxonomic identification of amplicon sequence variants and follows this tutorial: https://cran.r-project.org/web/packages/insect/vignettes/insect-vignette.html

load libraries
```{r}
library(insect)
```

read in the data (ASV table from DADA2)
make sure sample names are designated as row names, and not in their own column 
```{r}
seqs <- read.csv("/genetics/edna/workdir/GOApcod_2021/combined/trimmed/filtered/outputs/seqtab.csv", row.names = 1)
```

assign taxon IDs to the DADA2 output
```{r}
x <- char2dna(colnames(seqs))
## name the sequences sequentially
names(x) <- paste0("ASV", seq_along(x))
```

download the MiFish classifier from the insect tutorial locally (had to do this offsite) and then copy to the VM using the command prompt:  scp Downloads/classifier.rds kimberly.ledger@161.55.97.134:/home/kimberly.ledger/NBS_eDNA 
```{r}
classifier <- readRDS("/home/kimberly.ledger/NBS_eDNA/classifier.rds")
classifier
names(attributes(classifier))
```

classify 
* this takes around a minute to run 
```{r}
my_class_1_0 <- classify(x, classifier, threshold = 1)
```

output full classification table
```{r}
write.csv(my_class_1_0, "/genetics/edna/workdir/GOApcod_2021/combined/trimmed/filtered/outputs/asv_taxonomy_insect_1_0.csv")
```


output this table for comparison with blastn taxonomy 
```{r}
asv_insect <- my_class %>%
  select(representative, taxon, rank)

write.csv(asv_insect, "/genetics/edna/workdir/GOApcod_2021/combined/trimmed/filtered/outputs/asv_taxonomy_insect.csv")
```


if needed, remove column names that overwhelm the ASV table when printed... 
```{r}
colnames(seqs) <- NULL
```

transpose and bind the ASV table to the table of taxonomic information
```{r}
my_class_bind <- cbind(my_class, t(seqs))
```

Any sequences that return exact hits with at least one training sequence (or near matches if ping = 0.99 or similar) are assigned a score of NA

create an aggregated table that includes only one row for each unique taxon
```{r}
taxa <- aggregate(my_class_bind[3:12], my_class_bind["taxID"], head, 1)
counts <- aggregate(my_class_bind[13:ncol(my_class_bind)], my_class_bind["taxID"], sum)
my_class_short <- merge(taxa, counts, by = "taxID")
```

output taxonomic identification 
```{r}
write.csv(my_class_short, "/genetics/edna/workdir/bottomtrawl_2022/20230623/trimmed/filtered/outputs/taxon_table.csv")
```


