---
title: "filtering sequences"
author: "Kimberly Ledger"
date: "2023-01-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


DADA2 processing of trimmed reads (i.e. primer sequences already removed)

## access Rstudio on the VM 
* access the Rstudio server here: http://161.55.97.134:8787/
* Type in username and pw associated with VM

load packages
```{r, warning=FALSE}
library(dada2)
```

when running this code here are the only things that need to be customized
- file path 
- filter parameters and truncate lengths  
- merged sequence length filter 

name file path and check correct files are there - UPDATE FILE PATH DEPENDING ON DATASET - this should point to the trimmed folder
```{r}
path <- "/genetics/edna/workdir/bottomtrawl_2022/20230406/trimmed/"
head(list.files(path))
```

create lists for the forward and reverse fastq files 
```{r}
# Forward and reverse fastq filenames have format: SAMPLENAME-N_SX_R1.fastq and SAMPLENAME-N_SX_R2.fastq

fnFs <- sort(list.files(path, pattern="_R1.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2.fastq", full.names = TRUE))

# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
```

## inspect read quality 

visualize quality of forward reads 
```{r, warning=FALSE}
plotQualityProfile(fnFs[41:42])
```

visualize quality of reverse reads
```{r, warning=FALSE}
plotQualityProfile(fnRs[51:52])
```


## filter and trim 

assign filenames 
```{r}
# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered2", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered2", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```


**edit this code for trimming**
Use standard filtering parameters: maxN=0 (DADA2 requires no Ns), truncQ=2, rm.phix=TRUE and maxEE=2.

all gadid primer sequences usually: truncLen=c(120,120)
mifish and mideca sequences: truncLen=c(100,100) (using on first filtering of bottomtrawl_2022/2020406 seqs) -for filter2 i used 110 
for salmon cytB trying: truncLen=c(100,80)
```{r}
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(110,110),
              maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
              compress=TRUE, multithread=FALSE) # On Windows set multithread=FALSE
head(out)
```


## learn the error rates 
this takes a few minutes to run
```{r}
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)
```

plot
```{r}
plotErrors(errR, nominalQ=TRUE)
```

this shows the error rates for each possible nucleotide transition. the red lines are the expected error rates. just check that the estimated error rates (black lines) are a good fit to the observed rates (points). 

## subset to include filtFs and filtRs where files exist
```{r}
filtFs <- filtFs[file.exists(filtFs)]
filtRs <- filtRs[file.exists(filtRs)]
```


## apply the sample inference to the filtered and trimmed reads
```{r, results= "hide"}
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)
```

take a look at the output
```{r}
dadaFs[[1]]
```


## merge paired reads
```{r, warnings = FALSE, results= "hide"}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE, minOverlap = 8, maxMismatch = 2)
```


Inspect the merger data.frame from the first sample
```{r}
head(mergers[[1]])
```


## construct sequence table 

```{r}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```

```{r}
seqtab_test1 <- makeSequenceTable(mergers_test1)
dim(seqtab_test1)
```


Inspect distribution of sequence lengths
```{r}
table(nchar(getSequences(seqtab)))
```

Inspect distribution of sequence lengths
```{r}
table(nchar(getSequences(seqtab_test1)))
```

Remove sequences that are much longer or shorter than expected - UPDATE THE VALUES HERE BASED ON PRIMERSET
```{r}
#seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% 221] #gadidS1
#seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% 162] #gadidS2
#seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% 204] #gadidS3
#seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% 129] #gadidS4 and S5
#seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% 205] #gaididS6

#seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% c(151:154)] #salmon
#seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% c(153:184)] #mideca (amplicon size according to original publication)
seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% c(164:181)] #mifish
```

```{r}
seqtab_test1_1 <- seqtab_test1[,nchar(colnames(seqtab_test1)) %in% c(164:181)] #mifish
```

```{r}
seqtab_toobig <- seqtab_test1[,nchar(colnames(seqtab_test1)) %in% c(181:212)] #mifish
```


Remove chimeras
```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab2, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
sum(seqtab.nochim)/sum(seqtab2)
```

Remove chimeras
```{r}
seqtab.nochim_test <- removeBimeraDenovo(seqtab_toobig, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim_test)
sum(seqtab.nochim_test)/sum(seqtab_toobig)
```

Track reads through the pipeline
```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab2), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "correct_length","nonchim")
rownames(track) <- sample.names
head(track)
```

Track reads through the pipeline
```{r}
getN <- function(x) sum(getUniques(x))
track_test <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers_test1, getN), rowSums(seqtab_toobig), rowSums(seqtab.nochim_test))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track_test) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "correct_length","nonchim")
rownames(track_test) <- sample.names
head(track_test)
```



for S2, because S2-g05-S2-3-H had zero reads after filter output, the columns of track don't line up... for now it's faster for me to manually edit this than to figure out how to code... 
```{r}
#track[50,] <- c(0,0,0,0,0,0,0)
#track[51,] <- c(1,1,1,1,1,1,1)
```

for S5, because S2-g12-S5-3-H and S2-g12-S5-3-L had zero reads after filter output
```{r}
#track[50,] <- c(0,0,0,0,0,0,0)
#track[51,] <- c(0,0,0,0,0,0,0)
```


export dada2 filtering summary 
```{r}
out_path <- file.path(path, "filtered2", "outputs")

write.csv(track_test, file = file.path(out_path, "track_toobig.csv"))
```


**come back and re-evaluate read processing at a later time** there some big cuts of reads during merging and removing chimeras steps


## export for taxonomic identification
we will want to use the "seqtab.nochim" ASV table for taxonomic analyses

```{r}
write.csv(seqtab.nochim_test, file.path(out_path, "seqtab_toobig.csv"))
```


## also export the seqtab as a Fasta
```{r}
sq <- getSequences(seqtab.nochim_test)
id <- paste0("ASV", 1:length(sq))

library(seqinr)
write.fasta(sequences = as.list(sq), names=id, file.out = file.path(out_path, "myasvs_toobig.fasta"))
```



## and export the seqtab table with the numbered ASVs headers that correspond to those output in the FASTA file.
```{r}
# Make map between brief names and full sequences
briefToSeq <- colnames(seqtab.nochim_test)
names(briefToSeq) <- paste0("ASV", seq(ncol(seqtab.nochim_test))) # Seq1, Seq2, ...
# Make new sequence table with brief names
st.brief <- seqtab.nochim_test
colnames(st.brief) <- names(briefToSeq)

# export the seq table with brief names:
write.csv(st.brief, file= file.path(out_path, "ASVtable_toobig.csv"))
```


## create table of ASV labels and ASV sequences
```{r}
asv <- colnames(seqtab.nochim_test)
asv_id <- paste0("ASV", 1:length(asv))
asv_table <- data.frame(asv, asv_id)

write.csv(asv_table, file = file.path(out_path, "asv_id_table_toobig.csv"))
```

copied files to my local computer...
from command center (not logged into the VM): 

scp kimberly.ledger@161.55.97.134:/genetics/edna/workdir/gadids/20230221/S1_ND1/trimmed/filtered/outputs/* Downloads


